Spatial Reasoning and Statistical Graphics

### 2014-11-20 Paper/Draft1/TestingVisualAptitude.Rnw
calculating lineup percentage scores by section, with comment.
Lineup type scatterplot matrix, with visual and reasoning tests as well, plus some commentary.

\comment{Next few steps for the analysis: \\
(3) in terms of why we would use lineups in a test comparison ... lineups give us a way to say 'this is correct' which regular graphics don't. So we should put a spin sentence on at the intro and the back that we are suggesting to use lineups as a tool to not only evaluate individuals (which we can see is possible) but also as a tool to evaluate charts ... which we've done before ... and tie it into the regular testing literature ...\\
\textbf{A few sentences added, still need to tie into the existing literature}\\
(4) I'll take care of the intro to lineups\\
(5) Include lineup results from other studies. For that, we need to compare percent correct attempted data. 
}

@@

@@

From figure~\ref{fig:lineupblockpca} we see that again, PC1 accounts for general visual aptitude. 
% PC2 emphasizes the relationship between figure classification and lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for a joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.
\newtext{PC2 emphasizes the effect of performance on lineup test 1, which is associated with all of the measures of visual aptitude. PC3 emphasizes the association between performance on lineup test 3 and the visual search test, while PC4 emphasizes the association between performance on lineup test 2 and performance on the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings. }

\comment{If PC6-7 aren't important, then we still only have 5 PCs, or one additional PC from the grouped analysis. }
%\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.

>>>>

\comment{Next few steps for the analysis: \\
(3) in terms of why we would use lineups in a test comparison ... lineups give us a way to say 'this is correct' which regular graphics don't. So we should put a spin sentence on at the intro and the back that we are suggesting to use lineups as a tool to not only evaluate individuals (which we can see is possible) but also as a tool to evaluate charts ... which we've done before ... and tie it into the regular testing literature ...\\
\textbf{A few sentences added, still need to tie into the existing literature}\\
(4) I'll take care of the intro to lineups\\
(5) Include lineup results from other studies. For that, we need to compare percent correct attempted data. 
\textbf{In \texttt{lineup.section.pct}.}
}

@@

\newtext{Figure \ref{fig:lineuptaskmatrix} shows the correlations between the three lineup tasks and the figure classification, card rotation, and paper folding tasks. The visual search task was only slightly correlated with the three lineup tasks and has been omitted from this figure. Performance on lineup tasks 1 and 2, which dealt with the distributuion of two groups of numerical data, is most strongly correlated with the performance on the figure classification task, which measures general reasoning ability. Performance on lineup task 3, which dealt with the identification of nonnormal residual qq plots, is more associated with the card rotation and paper folding tests, which measure visuospatial ability. This suggests that certain lineup tasks may require more visual ability than others; in this case, the qq-plot lineups required participants to mentally rotate plots, compare vertical distances, and perform more mental manipulation than the first two lineup tasks. }

<<LineupTypeSPM,echo=FALSE,fig.pos="ht", include=T, eval=T, fig.width=8,fig.height=4, message=F, warning=F, fig.cap="Pairwise scatterplots of test scores, with lineups separated into the three lineup tasks. All lineup tasks are moderately correlated with the figure classification task, and while tasks 1 and 2 are most strongly correlated with figure classification, lineup task 3 is most strongly correlated with the card rotation and paper folding tasks. \\label{fig:lineuptaskmatrix}">>=
data <- lineup.section.summary[,c("lineup_1", "lineup_2", "lineup_3", "vis_search", "fig_class", "card_rot", "folding")]
grid <- expand.grid(x=1:ncol(data), y=1:ncol(data))
grid <- subset(grid, x != y & x<y)

all2 <- subset(all, grepl("lineup", yvar) & points & !grepl("vis_search", xvar))
names(all2)[1:4] <- c("yvar", "xvar", "y", "x")
all2mean <- rbind(
  ddply(subset(all2, !grepl("lineup", yvar)), .(xvar, yvar), summarize, label=paste0("Cor = ", round(cor(x,y), 3)), x=max(x), y=min(y)-1, hjust=1, vjust=0),
  ddply(subset(all2, grepl("lineup", xvar) & grepl("lineup", yvar)), .(xvar, yvar), function(df){
    with(df, data.frame(xvar=unique(yvar), yvar=unique(xvar), 
                        x=mean(range(y)), y=mean(range(x)), 
                        label=paste0("Correlation = \n", round(cor(x,y),3)),
                        hjust=.5, vjust=.5))
    })
)
par(mfrow=c(1,2),mar=c(4,4.5,.1,.1)

ggplot()+
  geom_point(data=subset(all2, points), aes(x=x, y=y)) + 
  geom_smooth(data=subset(all2, points), aes(x=x, y=y), method="lm") + 
  geom_text(data=all2mean, aes(x=x, y=y, label=label, hjust=hjust, vjust=vjust)) + 
  geom_line(data=subset(densities, grepl("lineup", xvar)), aes(x=x, y=y)) + 
  facet_grid(xvar~yvar, scales="free", labeller = labeller(xvar=fix.names, yvar=fix.names)) + 
  theme_bw() + 
  theme(axis.title=element_blank())
@

@@

From figure~\ref{fig:lineupblockpca} we see that again, PC1 accounts for general visual aptitude. 
% PC2 emphasizes the relationship between figure classification and lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for a joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.
\newtext{PC2 emphasizes the effect of performance on lineup test 1, which is associated with all of the measures of visual aptitude. PC3 emphasizes the association between performance on lineup test 3 and the visual search test, while PC4 emphasizes the association between performance on lineup test 2 and performance on the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings. }
% \comment{If PC6-7 aren't important, then we still only have 5 PCs, or one additional PC from the grouped analysis. }
%\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.
